---
title: "BDPROTO raw data sources aggregation script"
author: "Steven Moran <steven.moran@uzh.ch>"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
  pandoc_args: --webtex
---

# Overview

This a script to aggregate the raw data sources in BDPROTO into single CSV and RData files.

In BDPROTO versions 1.0 and 1.1, we integreated separate metadata for each raw data source (`bdproto-original`, `uz`, `huji`, and `ancient-near-east`). To make things easier to maintain and process, we extracted the metadata fields pertinent to all source inventories into a separate CSV file, which we then aggregated into the combined raw data sources.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(stringi)
library(knitr)
library(testthat)
```


# Metadata

Read in the metadata.

```{r, warning=FALSE, message=FALSE}
metadata <- read_csv('BDPROTO metadata - bdproto_metadata.csv')
```

Let's do some tests to make sure things like Glottolog glottocodes are valid. First load the Glottolog's languiods data.

```{r, warning=FALSE, message=FALSE}
glottolog <- read_csv('glottolog_languoid.csv/languoid.csv')
```

Then make sure all BDPROTO glottocodes are represented in that file. Note that we have a bunch of NAs (we know there exists no glottocode for these languages). These should all be NA.

```{r}
metadata$Glottocode[which(!(metadata$Glottocode %in% glottolog$id))]
```

Let's eyeball the canonical language names for typos.

```{r}
table(metadata$LanguageName) %>% kable() %>% head()
```

And the same for the language family names.

```{r}
table(metadata$LanguageFamily) %>% kable() %>% head()
```

And the same for the root level language families.

```{r}
table(metadata$LanguageFamilyRoot) %>% kable() %>% head()
```


# Load and clean raw data sources

## Original BDPROTO inventories 

Load the data.

```{r, warning=FALSE, message=FALSE}
bdproto.inventories <- read_csv('bdproto-original/bdproto-inventories.csv')
expect_equal(nrow(bdproto.inventories), 2862)
```


## UZ inventories

Load the inventories from the `UZ` source.

```{r, warning=FALSE, message=FALSE}
uz.raw <- read_csv("uz/uz-inventories.csv")
```


## Ancient-near-east inventories

Load the ANE data.

```{r, warning=FALSE, message=FALSE}
ane.raw <- read_csv('ancient-near-east/AnNeEa-inventories.csv')
```

It is a slightly different format. It has blank rows between inventories and the field for IDs and names do not repeat. Let's remove or fill those in.

```{r}
dim(ane.raw[rowSums(is.na(ane.raw)) == ncol(ane.raw), ])
ane.raw <- ane.raw[rowSums(is.na(ane.raw)) != ncol(ane.raw), ]
rownames(ane.raw) <- NULL
```

This should be zero.

```{r}
expect_equal(nrow(ane.raw[rowSums(is.na(ane.raw)) == ncol(ane.raw), ]), 0)
```

Fill in the blank cells.

```{r}
ane.raw <- ane.raw %>% fill(BdprotoID)
ane.raw <- ane.raw %>% fill(SourceLanguageName)
ane.raw <- ane.raw %>% fill(SpecificDialect)
```

## HUJI Inventories 

```{r, warning=FALSE, message=FALSE}
huji.inventories <- read_csv('huji/BDPROTO Jerusalem - Inventories.csv')
```

Check the inventories for blank rows -- if yes, they need to be removed so not to introduce NA as a phoneme.

```{r}
expect_equal(nrow(huji.inventories[rowSums(is.na(huji.inventories)) == ncol(huji.inventories), ]), 0)
```

Check that all huji inventories are represented in the metadata file.

```{r}
expect_false(any(!(huji.inventories$BdprotoID %in% metadata$BdprotoID)))
```

Make sure there are no duplicate phoneme values in each language.

```{r}
expect_equal(nrow(huji.inventories %>% group_by(BdprotoID, Phoneme) %>% select(BdprotoID, SourceLanguageName, Phoneme) %>% filter(n()>1) %>% arrange(BdprotoID, Phoneme)), 0)

# If so, these are the ones
# huji.inventories %>% group_by(BdprotoID, Phoneme) %>% select(BdprotoID, SourceLanguageName, Phoneme) %>% filter(n()>1) %>% arrange(BdprotoID, Phoneme)
```

# Merge the data sources

Merge the three inventory data sources together.

First check that the input data sources haven't changed in size or if something has been lost along the way.

```{r}
expect_equal(nrow(bdproto.inventories), 2862)
expect_equal(nrow(uz.raw), 479)
expect_equal(nrow(ane.raw), 666)
expect_equal(nrow(huji.inventories), 3954)
```

Now we stack the data frames.

```{r}
inventories <- bind_rows(bdproto.inventories, uz.raw, ane.raw, huji.inventories)
inventories <- ungroup(inventories)
rownames(inventories) <- NULL
```

The concatenation should reflect the same number of rows combined above.

```{r}
expect_equal(nrow(inventories), (nrow(bdproto.inventories) + nrow(uz.raw) + nrow(ane.raw) + nrow(huji.inventories)))
```

And merge in the metadata.

```{r}
results <- left_join(inventories, metadata, by = c("BdprotoID" = "BdprotoID", "SourceLanguageName" = "SourceLanguageName"))
```

How's it look?

```{r}
results %>% head() %>% kable()
```

Let's run some checks.

Anything weird about the phoneme counts? Recall, some inventories are consonants or vowels only.

```{r}
results %>% group_by(BdprotoID, InventoryType) %>% summarize(segments = n()) %>% arrange(segments)
temp <- results %>% select(BdprotoID, InventoryType) %>% distinct()
table(temp$InventoryType, exclude = FALSE)
```

There should be no duplicate phonemes in the merged data set (within each language).

```{r}
results %>% select(BdprotoID, Phoneme) %>% group_by(BdprotoID, Phoneme) %>% distinct() %>% arrange(Phoneme) %>% filter(n() > 1)
```

Check BDPROTO segments against PHOIBLE.

Load PHOIBLE.

```{r}
url_ <- "https://github.com/phoible/dev/blob/master/data/phoible.csv?raw=true"
col_types <- cols(InventoryID='i', Marginal='l', .default='c')
phoible <- read_csv(url(url_), col_types=col_types)
```

Which segments in BDPROTO are not in PHOIBLE?

```{r}
bdproto.segments <- results %>% select(Source, Phoneme) %>% group_by(Source, Phoneme) %>% summarize(count=n())
table(bdproto.segments$Phoneme %in% phoible$Phoneme)
temp <- bdproto.segments[which(!(bdproto.segments$Phoneme %in% phoible$Phoneme)),]
write_csv(temp, 'missing-phonemes.csv')
```

How about if we NDF the characters?

```{r}
results$PhonemeNFD <- stri_trans_nfd(results$Phoneme)
bdproto.segments <- results %>% select(Source, PhonemeNFD) %>% group_by(Source, PhonemeNFD) %>% summarize(count=n())
table(bdproto.segments$PhonemeNFD %in% phoible$Phoneme)
temp <- bdproto.segments[which(!(bdproto.segments$PhonemeNFD %in% phoible$Phoneme)),]
write_csv(temp, 'inventories_NFD.csv')
```

```{r}
results$PhonemeNFD <- stri_trans_nfd(results$Phoneme)
results$PhonemeNFD <- stri_trans_nfc(results$Phoneme)
bdproto.segments <- results %>% select(Source, PhonemeNFD) %>% group_by(Source, PhonemeNFD) %>% summarize(count=n())
phoible$PhonemeNFD <- stri_trans_nfd(phoible$Phoneme)
phoible$PhonemeNFD <- stri_trans_nfc(phoible$Phoneme)

table(bdproto.segments$PhonemeNFD %in% phoible$Phoneme)
temp <- bdproto.segments[which(!(bdproto.segments$PhonemeNFD %in% phoible$Phoneme)),]
write_csv(temp, 'inventories_NFD.csv')
```

# 



Write to disk.

```{r}
write_csv(inventories, 'inventories.csv')

```


## Add the PHOIBLE segment features

```{r}
load(url('https://github.com/phoible/dev/blob/master/data/phoible.RData?raw=true'))
features <- phoible %>% select(Phoneme,tone,stress,syllabic,short,long,consonantal,sonorant,continuant,delayedRelease,approximant,tap,trill,nasal,lateral,labial,round ,labiodental,coronal,anterior,distributed,strident,dorsal,high,low,front,back,tense,retractedTongueRoot,advancedTongueRoot,periodicGlottalSource,epilaryngealSource,spreadGlottis,constrictedGlottis,fortis,raisedLarynxEjective,loweredLarynxImplosive,click) %>% distinct()
expect_equal(nrow(features), 3169)
num_inventories <- nrow(inventories)

# Merge
inventories <- left_join(inventories, features)
expect_equal(nrow(inventories), num_inventories)
rm(num_inventories, phoible, features)

# Identify potential coding errors in HUJI
bad.chars <- inventories %>% filter(Source=="HUJI") %>% group_by(Phoneme, consonantal) %>% select(Phoneme, consonantal) %>% filter(is.na(consonantal)) %>% distinct()
write.csv(bad.chars, file='bad-chars.csv')
rm(bad.chars)

# TODO: identify potential coding errors in other resources
```

## Identify consonant vs vowels

```{r}
phonemes <- inventories %>% group_by(Phoneme) %>% select(Phoneme, consonantal, syllabic) %>% distinct()
table(phonemes$consonantal, exclude=F) # 319 NAs

# [consonantal] is a proxy for consonants and does not work in all cases, see below
phonemes.cs <- phonemes %>% summarize(is.consonantal = ifelse(consonantal=="+", TRUE, FALSE))
phonemes.vs <- phonemes %>% summarize(is.not.consonantal = ifelse(consonantal=="-" & syllabic=="+", TRUE, FALSE))
phonemes.status <- left_join(phonemes.cs, phonemes.vs)
phonemes.status[which(phonemes.status$is.consonantal == phonemes.status$is.not.consonantal),]

# We write the data to disk and then fix the missing data points by hand
# To this we add a `type` column (copied from `is.consonantal` and updated by hand)
# write.csv(phonemes.status, file='phonemes-status.csv')
rm(phonemes, phonemes.cs, phonemes.vs, phonemes.status)

# TODO: when the input data sources are changed or updated in the future, we should:
# read in the current output csv `phonemes-status-by-hand.csv` and check newly inputted 
# segments against it and then add correct them.

# The hand curated consonant vs vowel types
segment.types <- read.csv('phonemes-status-by-hand.csv', header=TRUE, stringsAsFactors=FALSE)
glimpse(segment.types)
glimpse(inventories)

# Merge in the segment type information
inventories <- left_join(inventories, segment.types)
table(inventories$type, exclude=F) # Looks like 16 NA segments to deal with

rm(segment.types)
```

# Data checks

TODO: 

* check that all Glottocodes are valid
* remove duplicate in Phoneme

* 
```{r}
# x <- inventories %>% select(BdprotoID, Glottocode) %>% distinct() %>% filter(!is.na(Glottocode)) %>% filter(Glottocode!="")
# x$Glottocode
# grepl("")

# InventoryType introduces NA during the merge. Make all non-values NA.
inventories %>% select(InventoryType)
table(inventories$InventoryType, exclude = F)
inventories$InventoryType[inventories$InventoryType==""] <- NA
table(inventories$InventoryType, exclude = F)
```

## Identify how many unique data points

```{r}
# Identify the unique entries given their IDs, language names, and reported Glottocodes
# We write these to disk and then update them by hand

# inventories-duplicates <- inventories %>% select(BdprotoID, LanguageName, Glottocode) %>% unique() %>% arrange(LanguageName)
# write.csv(x, 'inventories-duplicates.csv')

duplicate.entries <- read.csv('inventories-duplicates-by-hand.csv', header=T, stringsAsFactors=F)
inventories <- left_join(inventories, duplicate.entries)
rm(duplicate.entries)
```

# Save 

```{r}
save(inventories, file='../bdproto.Rdata')
write.csv(inventories, file='../bdproto.csv', row.names = FALSE)
```

