---
title: "BDPROTO descriptive stats"
author: "Steven Moran"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
  pandoc_args: --webtex
---

Some preliminary descriptive stats about the current BDPROTO data.
  
```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)
```

Load BDPROTO data.

```{r}
load(file='../bdproto.Rdata')
```

Display a subset of the data.

```{r}
kable(inventories %>% select(BdprotoID, LanguageName, Glottocode, Phoneme, LanguageFamily, TimeDepthYBP) %>% arrange(BdprotoID)) %>% head()
```

How many total inventories are there (with duplicate languages)?

```{r}
nrow(inventories %>% select(BdprotoID) %>% unique())
```

How many data points are there per data source?

```{r}
sources <- inventories %>% select(BdprotoID, LanguageName, Source) %>% 
  group_by(BdprotoID, LanguageName, Source) %>% distinct()
table(sources$Source)
```

How many unique phonemes are there?

```{r}
# TODO: some of these will have to be cleaned up, so the figure will be lower.
nrow(inventories %>% select(Phoneme) %>% group_by(Phoneme) %>% distinct())
```

How many language families are tagged?

```{r}
nrow(inventories %>% select(LanguageFamily) %>% group_by(LanguageFamily) %>% distinct())
```

What are they?

```{r}
table(inventories %>% select(LanguageFamily) %>% group_by(LanguageFamily) %>% distinct()) %>% kable()
```

How many distinct vs duplicate data points are there?

```{r}
temp <- inventories %>% select(BdprotoID, LanguageName, Glottocode, Duplicate) %>% unique() %>% arrange(LanguageName)
table(temp$Duplicate)
```

That is, there are 178 independent data points and 79 duplicated ones. Through manual inspection, we identified 3 sets of duplicates (9 data points), and the rest are doubles. This means: 70/2=25 + 3 + 178 == **206 unique language data points**.

How many distinct Glottocodes are there? Some are NA (i.e. don't exist, e.g. Altaic) and others have not yet been identified.

```{r}
bdproto.glottocodes <- inventories %>% select(Glottocode) %>% filter(!is.na(Glottocode)) %>% unique()
nrow(bdproto.glottocodes) # 187 without NAs or ""
```

How many Glottocodes are NA (i.e. we know there exists no Glottocode at the moment)?

```{r}
# inventories %>% group_by(BdprotoID, Glottocode) %>% select(BdprotoID, Glottocode) %>% distinct() %>% filter(n()<1)
temp <- as.data.frame(inventories %>% select(BdprotoID, Glottocode) %>% distinct())
table(temp$Glottocode=="") # 22
table(temp$Glottocode=="", exclude=F) # 11 NA

# TODO:
# library(testthat)
# expect_that() # 224 + 22 + 11 = 257
```

Get the Glottolog family IDs (this doesn't do much because language families don't have latitude, longitude, area, etc.)

```{r}
glottolog <- read.csv('glottolog_languoid.csv/languoid.csv', header=T, stringsAsFactors = F)
glottlog.families <- glottolog %>% select(family_id) %>% distinct()
```

How many BDPROTO Glottocodes are in the Glottolog top-level language family trees?

```{r}
table(bdproto.glottocodes$Glottocode %in% glottlog.families$family_id)
```

Which ones?

```{r}
kable(bdproto.glottocodes[which(bdproto.glottocodes$Glottocode %in% glottlog.families$family_id), ])
```

How many segments do the proto-languages have? (Note some inventories only have consonant or vowel inventories specified, so we filter by the `InventoryType` field.)

```{r}
inventories.counts <- inventories %>% select(BdprotoID, LanguageName, InventoryType) %>% group_by(BdprotoID, LanguageName, InventoryType) %>% summarize(segments=n()) %>% arrange(segments)
dim(inventories.counts)
```

Drop consonant and vowel only inventories (currently 10 data points)

```{r}
inventories.counts.cs.vs <- inventories.counts %>% filter(is.na(InventoryType))
dim(inventories.counts.cs.vs)
```

What is the median and mean number of segments in the sample (for full consonant and vowel inventories)?

```{r}
summary(inventories.counts.cs.vs$segments)
```

All segment types.

```{r}
# TODO: fix the NAs in the input data
table(inventories$type, exclude=F)
```

Get consonant counts and stats of ALL inventories.

```{r}
inventories.consonants <- inventories %>% filter(is.na(InventoryType) | InventoryType=="consonants") %>% filter(Source!="ANE") 
table(inventories.consonants$InventoryType, exclude=F)
```

Get the consonant counts per inventory (252 data points).

```{r}
c.counts <- inventories.consonants %>% select(BdprotoID, Phoneme, type) %>% filter(type=="C") %>%  group_by(BdprotoID) %>% summarize(consonants = n())
summary(c.counts$consonants)
```

Get consonant counts and stats of original BDPROTO (one data point per genealogical unit).

```{r}
og.bdproto.cs <- inventories %>% filter(is.na(InventoryType) | InventoryType=="consonants") %>% filter(Source=="BDPROTO")
```

Get the consonant counts per inventory (252 data points).

```{r}
og.c.counts <- og.bdproto.cs %>% select(BdprotoID, Phoneme, type) %>% filter(type=="C") %>% group_by(BdprotoID) %>% summarize(consonants = n())
summary(og.c.counts$consonants)
```

Get vowel counts and stats of ALL inventories.

```{r}
inventories.vowels <- inventories %>% filter(is.na(InventoryType) | InventoryType=="vowels") %>% filter(Source!="ANE")
table(inventories.vowels$InventoryType, exclude=F)
```

Get the vowel counts per inventory (should be 252 data points).

```{r}
v.counts <- inventories.vowels %>% select(BdprotoID, Phoneme, type) %>% filter(type=="V") %>% group_by(BdprotoID) %>% summarize(vowels = n())
summary(v.counts$vowels)
```

Get vowel counts and stats of original BDPROTO (one data point per genealogical unit).

```{r}
og.bdproto.vs <- inventories %>% filter(is.na(InventoryType) | InventoryType=="vowels") %>% filter(Source=="BDPROTO")
```

Get the consonant counts per inventory (252 data points).

```{r}
og.v.counts <- og.bdproto.vs %>% select(BdprotoID, Phoneme, type) %>% filter(type=="V") %>% group_by(BdprotoID) %>% summarize(vowels = n())
summary(og.v.counts$vowels)
```

Plot segment counts (all inventories).

```{r}
inventories.counts$BdprotoID <- factor(inventories.counts$BdprotoID, levels=inventories.counts$BdprotoID[order(-inventories.counts$segments)])
qplot(inventories.counts$LanguageName, inventories.counts$segments)
```

What is the frequency of segments across the proto-languages? Use only inventories that have both consonant and vowel descriptions.

```{r}
inventories.cs.vs <- inventories %>% filter(is.na(InventoryType))
segment.counts <- inventories.cs.vs %>% select(Phoneme) %>% group_by(Phoneme) %>% summarize(count=n()) %>% arrange(desc(count)) %>% filter(!is.na(Phoneme))
dim(segment.counts)
head(segment.counts)
```

Get percentages.

```{r}
total.inventories <- nrow(inventories.cs.vs %>% select(BdprotoID) %>% distinct())
segment.counts$Percentage <- segment.counts$count/total.inventories
```

Plot it.

```{r}
segment.counts$Phoneme <- factor(segment.counts$Phoneme, levels=segment.counts$Phoneme[order(-segment.counts$count)])
ggplot(segment.counts, aes(x=Phoneme, y=count))+
geom_point() + 
ylab('count') +
xlab('Phoneme')
```

Plot just the top 50 most frequent segments.

```{r}
top <- head(segment.counts, n=50)
# qplot(top$Phoneme, top$count)
qplot(top$Phoneme, top$Percentage)
```

Get phoible phonemes for comparison.

```{r}
phoible <- read_csv('https://raw.githubusercontent.com/phoible/dev/master/data/phoible.csv')

# PHOIBLE inventories are not unique (3020)
num.phoible.inventories <- nrow(phoible %>% select(InventoryID) %>% distinct())

# Get number of unique inventories by Glottocode (2185)
num.phoible.inventories <- nrow(phoible %>% select(Glottocode) %>% distinct())

# Get phoneme counts and percentages
phoible.phonemes <- phoible %>% select(Glottocode, Phoneme) %>% group_by(Glottocode) %>% distinct() %>% group_by(Phoneme) %>% summarize(count=n()) %>% arrange(desc(count))

phoible.phonemes$Percentage <- phoible.phonemes$count/num.phoible.inventories
head(phoible.phonemes)

phoible.phonemes$Phoneme <- factor(phoible.phonemes$Phoneme, levels=phoible.phonemes$Phoneme[order(-phoible.phonemes$count)])
```

Segment counts (disregards genealogical relatedness and areal proximity).

```{r}
phoible.counts <- phoible %>% select(InventoryID, Phoneme, SegmentClass) %>% group_by(InventoryID) %>% summarize(segments=n())
summary(phoible.counts$segments)
```

Consonant counts (disregards genealogical relatedness and areal proximity).

```{r}
phoible.cs <- phoible %>% select(InventoryID, Phoneme, SegmentClass) %>% filter(SegmentClass == "consonant") %>% group_by(InventoryID) %>% summarize(segments=n())
summary(phoible.cs$segments)
```

Vowel counts (disregards genealogical relatedness and areal proximity).

```{r}
phoible.vs <- phoible %>% select(InventoryID, Phoneme, SegmentClass) %>% filter(SegmentClass == "vowel") %>% group_by(InventoryID) %>% summarize(segments=n())
summary(phoible.vs$segments)
```

Tone counts (disregards genealogical relatedness and areal proximity).

```{r}
phoible.ts <- phoible %>% select(InventoryID, Phoneme, SegmentClass) %>% filter(SegmentClass == "tone") %>% group_by(InventoryID) %>% summarize(segments=n())
summary(phoible.ts$segments)
```

Frequency distribution of phonemes in phoible.

```{r}
qplot(phoible.phonemes$Phoneme, phoible.phonemes$Percentage)
```

Frequency distribution of 50 most frequent phonemes in phoible.

```{r}
top.phoible <- head(phoible.phonemes, n=50)
qplot(top.phoible$Phoneme, top$Percentage)
```

Combine the phoneme counts to plot them together in one graph.

```{r}
x <- phoible.phonemes
colnames(x) <- c("Phoneme", "Phoible.count", "Phoible.percentage")
y <- segment.counts
colnames(y) <- c("Phoneme", "Bdproto.count", "Bdproto.percentage")
z <- left_join(x,y)
rm(x,y)

# Reorder phonemes as factor
z$Phoneme <- factor(z$Phoneme, levels=z$Phoneme[order(-z$Phoible.count)])
```

Try top 30 phonemes.

```{r}
top.z <- head(z, n=30)

ggplot(data = top.z, aes(x = Phoneme, group=1)) +
  geom_line(aes(y = Phoible.percentage, color = "PHOIBLE")) + 
  geom_line(aes(y = Bdproto.percentage, color = "BDPROTO")) +
  ylab("Percentage of language sample") +
  xlab("Most frequent phonemes reported in PHOIBLE") +
  labs(color="Database") +
  theme_bw()
```

Try top 30 phonemes without color.

```{r}
top.z <- head(z, n=30)

ggplot(data = top.z, aes(x = Phoneme, group=1)) +
  geom_line(aes(y = Bdproto.percentage, linetype = "BDPROTO")) +
  geom_line(aes(y = Phoible.percentage, linetype = "PHOIBLE")) + 
  ylab("Percentage of language sample") +
  xlab("Most frequent phonemes reported in PHOIBLE and BDPROTO") +
  labs(linetype="Database") +
  theme_bw()
```

Try top 50 phonemes.

```{r}
top.z <- head(z, n=50)

# Some phonemes occur very rarely in BDPROTO (or not at all) compared to PHOIBLE, e.g. /ts/ occurs at 2.5%. High tone not at all.
top.z %>% filter(Phoneme=="ts")
top.z %>% filter(Phoneme=="˦")

ggplot(data = top.z, aes(x = Phoneme, group=1)) +
  geom_line(aes(y = Phoible.percentage, color = "PHOIBLE")) + 
  geom_line(aes(y = Bdproto.percentage, color = "BDPROTO")) +
  ylab("Percentage of language sample") +
  xlab("Most frequent phonemes reported in PHOIBLE") +
  labs(color="Database") +
  theme_bw()
```

Get dates.

```{r}
inventories.dates <- inventories.cs.vs %>% select(BdprotoID, LanguageName, LanguageFamily, TimeDepthYBP) %>% group_by(BdprotoID, LanguageName, LanguageFamily, TimeDepthYBP) %>% distinct() %>% arrange(desc(TimeDepthYBP)) %>% filter(TimeDepthYBP < 10001)
head(inventories.dates)
```

Get coverage by Glottolog macroarea.

```{r}
# Get the geo/genealogical data from Glottolog
geo <- read.csv(url("https://cdstar.shh.mpg.de/bitstreams/EAEA0-E7DE-FA06-8817-0/languages_and_dialects_geo.csv"), stringsAsFactors = FALSE)

# Merge with the BDPROTO data points
temp <- left_join(bdproto.glottocodes, geo, by=c("Glottocode"="glottocode"))
head(temp)

# Problem here is that language family level codes, e.g. grea1284 (Greater Central Philippine), are not associated with macroarea, lat, long, etc.

# So this number is completely off and represent essentially language isolates (which are their own family)
table(temp$macroarea, exclude=FALSE)

# TODO: infer the macroarea of a language family in Glottolog by it's daughter language(s)
```


```{r}
# Be kind and clean up the workspace
rm(list = ls())
```

<!-- 
########################################################
# Merge results in segments with NA for feature values #
########################################################
inventories %>% dplyr::filter(is.na(consonantal)) %>% select(ID, Name, Glottocode) %>% unique()
missing.segments <- inventories %>% filter(is.na(consonantal)) %>% select(Phoneme) %>% unique()
write.table(missing.segments, "missing-segments.csv", sep="\t", quote = F, row.names = F)

# There are 36 languages that have one or more missing feature vectors.
dim(inventories %>% filter(is.na(consonantal)) %>% select(ID, Name) %>% unique())
no.feature.vectors <- anti_join(inventories, features)
no.feature.vectors <- no.feature.vectors %>% select(ID, Name, Phoneme, Source) %>% arrange(Name) %>% arrange(ID)
nrow(no.feature.vectors) # 91 feature vectors missing (with duplicates)
dim(no.feature.vectors %>% distinct(LanguageName)) # 36 distinct segments
head(no.feature.vectors)
-->