---
title: "BDPROTO aggregation script"
author: "Steven Moran\n"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
  pandoc_args: --webtex
---

# Overview

This script aggregates the raw data sources in BDPROTO into single CSV file.

In BDPROTO versions 1.0 and 1.1, we integrated separate metadata for each raw data source (`bdproto-original`, `uz`, `huji`, and `ancient-near-east`). To make things easier to maintain and process, we extracted the metadata fields pertinent to all source inventories into a separate CSV file, which we then aggregated into the combined raw data sources.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(stringi)
library(knitr)
library(testthat)
```


# Metadata

Read in the metadata. It's stored in a Google sheet for editors.

```{r, warning=FALSE, message=FALSE}
metadata <- read_csv('BDPROTO metadata - bdproto_metadata.csv')
```

Let's do some tests to make sure things like Glottolog glottocodes are valid. First load the Glottolog's languiods data.

```{r, warning=FALSE, message=FALSE}
glottolog <- read_csv('glottolog_languoid.csv/languoid.csv')
```

Then make sure all BDPROTO glottocodes are represented in that file. Note that we have a bunch of NAs (we know there exists no glottocode for these languages). These should all be NA.

```{r}
metadata$Glottocode[which(!(metadata$Glottocode %in% glottolog$id))]
```

There should be no NAs in several fields in the metadata.

```{r}
expect_false(any(is.na(metadata$LanguageName)))
expect_false(any(is.na(metadata$LanguageFamily)))
expect_false(any(is.na(metadata$LanguageFamilyRoot)))
```

We can uncomment this code to eyeball the canonical language names, families, and roots for typos.

```{r}
# table(metadata$LanguageName) %>% kable()
# table(metadata$LanguageFamily) %>% kable()
# table(metadata$LanguageFamilyRoot) %>% kable()
```


# Load and merge raw data sources

Load the original bdproto data.

```{r, warning=FALSE, message=FALSE}
bdproto <- read_csv('bdproto-original/bdproto-inventories.csv')
expect_equal(nrow(bdproto), 2862)
```

Load the inventories from the `UZ` source.

```{r, warning=FALSE, message=FALSE}
uz <- read_csv("uz/uz-inventories.csv")
expect_equal(nrow(uz), 424)
```

Load the ancient-near-east (ANE) inventories

```{r, warning=FALSE, message=FALSE}
ane <- read_csv('ancient-near-east/AnNeEa-inventories.csv')
expect_equal(nrow(ane), 684)
```

ANE is a slightly different format. It has blank rows between inventories and the field for IDs and names do not repeat. Let's remove or fill those in.

```{r}
dim(ane[rowSums(is.na(ane)) == ncol(ane), ])
ane <- ane[rowSums(is.na(ane)) != ncol(ane), ]
rownames(ane) <- NULL
```

This should be zero.

```{r}
expect_equal(nrow(ane[rowSums(is.na(ane)) == ncol(ane), ]), 0)
```

Fill in the blank ANE cells.

```{r}
ane <- ane %>% fill(BdprotoID)
ane <- ane %>% fill(SourceLanguageName)
ane <- ane %>% fill(SpecificDialect)
```

The result has less rows.

```{r}
expect_equal(nrow(ane), 664)
```

Load the HUJI Inventories.

```{r, warning=FALSE, message=FALSE}
huji <- read_csv('huji/BDPROTO Jerusalem - Inventories.csv')
expect_equal(nrow(huji), 3829)
```

Merge the three inventory data sources together.

```{r}
all_sources <- bind_rows(bdproto, uz, ane, huji) %>% ungroup()
rownames(all_sources) <- NULL
```

The concatenation should reflect the same number of rows combined above.

```{r}
expect_equal(nrow(all_sources), (nrow(bdproto) + nrow(uz) + nrow(ane) + nrow(huji)))
```

Make sure everything important is in the inventories and the metadata before merging them.

```{r}
expect_true(all(all_sources$BdprotoID %in% metadata$BdprotoID))
expect_true(all(metadata$BdprotoID %in% all_sources$BdprotoID))
expect_true(all(all_sources$SourceLanguageName %in% metadata$SourceLanguageName))
expect_true(all(metadata$SourceLanguageName %in% all_sources$SourceLanguageName))

# If something fails, adapt this code to find out what:
# metadata[which(!(metadata$BdprotoID %in% all_sources$BdprotoID)),]
```

And merge in the metadata.

```{r}
inventories <- left_join(all_sources, metadata, by = c("BdprotoID" = "BdprotoID", "SourceLanguageName" = "SourceLanguageName"))
```

How's it look?

```{r}
inventories %>% head() %>% kable()
```

Let's run some checks. No NAs.

```{r}
expect_false(any(is.na(inventories$BdprotoID)))
expect_false(any(is.na(inventories$SourceLanguageName)))
expect_false(any(is.na(inventories$LanguageName)))
expect_false(any(is.na(inventories$LanguageFamily)))
expect_false(any(is.na(inventories$LanguageFamilyRoot)))
expect_false(any(is.na(inventories$InventoryType)))
expect_false(any(is.na(inventories$Type)))
expect_false(any(is.na(inventories$Phoneme)))
# Todo
# expect_false(any(is.na(inventories$BibtexKey)))
```

Make sure there are no duplicate phoneme values within each language.

```{r}
expect_equal(nrow(inventories %>% group_by(BdprotoID, Phoneme) %>% select(BdprotoID, SourceLanguageName, Phoneme) %>% filter(n()>1) %>% arrange(BdprotoID, Phoneme)), 0)

# If so, these are the ones:
# inventories %>% group_by(BdprotoID, Phoneme) %>% select(BdprotoID, SourceLanguageName, Phoneme) %>% filter(n()>1) %>% arrange(BdprotoID, Phoneme)
```

Make sure there are no empty `Phoneme` rows.

```{r}
expect_equal(nrow(inventories %>% filter(Phoneme=="")), 0)
```

Are there still parentheses in the `Phoneme` column? There should not be.

```{r}
inventories %>% select(Phoneme) %>% filter(grepl('\\(', Phoneme)) %>% distinct()
```

Anything weird about the phoneme counts? Recall, some inventories are consonants or vowels only.

```{r}
inventories %>% group_by(BdprotoID, InventoryType) %>% summarize(segments = n()) %>% arrange(segments) %>% head() %>% kable()

inventories %>% group_by(BdprotoID, InventoryType) %>% summarize(segments = n()) %>% arrange(segments) %>% tail() %>% kable()
```

What the distribution of inventory types?

```{r}
temp <- inventories %>% select(BdprotoID, InventoryType) %>% distinct()
table(temp$InventoryType, exclude = FALSE)
```

Check BDPROTO segments against PHOIBLE. First load PHOIBLE.

```{r}
col_types <- cols(InventoryID='i', Marginal='l', .default='c')
phoible <- read_csv(url('https://github.com/phoible/dev/blob/master/data/phoible.csv?raw=true'), col_types=col_types)
```

Which segments in BDPROTO are not in PHOIBLE?

```{r}
bdproto.segments <- inventories %>% select(Source, LanguageName, Phoneme) %>% group_by(Source, LanguageName, Phoneme) %>% summarize(count=n())
table(bdproto.segments$Phoneme %in% phoible$Phoneme)
temp <- bdproto.segments[which(!(bdproto.segments$Phoneme %in% phoible$Phoneme)),]
# write_csv(temp, 'missing-phonemes-by-lg.csv')
```

How about if we NDF the characters?

```{r}
inventories$PhonemeNFD <- stri_trans_nfd(inventories$Phoneme)
bdproto.segments <- inventories %>% select(Source, PhonemeNFD) %>% group_by(Source, PhonemeNFD) %>% summarize(count=n())
table(bdproto.segments$PhonemeNFD %in% phoible$Phoneme)
temp <- bdproto.segments[which(!(bdproto.segments$PhonemeNFD %in% phoible$Phoneme)),]
# write_csv(temp, 'inventories_NFD.csv')
```

Let's write the merged results to disk.

```{r}
write_csv(inventories, '../bdproto.csv')
```

